ODGI: scalable tools for pangenome graphs

### 2021-05-31 manuscript/main.tex
shorter intro

    A pangenome models the full set of genomic elements in a given species or clade~\citep{32453966}. Indeed, these
    data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
    approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
    genetic variation across a population, but graph-based data structures present computational overheads: in addition
    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
    graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
    efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions,
    such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs which present paths with
    very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the
    paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the
    nodes they cross.

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
    implementations would require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
    efficiently scaling up to hundreds of large vertebrate genomes.

>>>>

    A pangenome models the full set of genomic elements in a population~\citep{32453966}. Indeed, these data structures
    encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches
    which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
    genetic variation across a population, but graph-based data structures present computational overheads: in addition
    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
    graph. Furthermore, the changes must perform efficiently on the complex graphs generated from whole-genome
    alignment of vertebrates. Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric
    chromosomes lead to graphs which present very high-depth nodes. The node depth is defined as the
    number of times in which the node is crossed by all the paths embedded in the graph. High depth increases the
    complexity of the operations performed on the paths and the nodes they cross.

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
    index-free manner, avoiding the burning task to create an index structure. Indeed, index-constrained
    implementations require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
    format) in parallel, which is one of the major bottlenecks when working with very large and deep graphs,
    scaling up to hundreds of large vertebrate genomes.
